Vagueness, Noise, and Communication

### 2014-09-18 paper/03_simulations.tex
more data exploration with varying tolerance and impairment

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \alpha) = \expo \left ( -
    \frac{\card{\mystate{i} - \mystate{j}}^s}{ \alpha^2} \right ) \,,
\end{align}
where $\alpha$ is an imprecision parameter. When $\alpha=0$ agents
perfectly discriminate between states; when $\alpha \rightarrow
\infty$ agents cannot discriminate states at
all. Figure~\ref{fig:NosofskySim} gives an impression of
Nosofsky-similarity for different parameter values. Other
formalizations of perceptual similarity are possible, including ones
that allow for different discriminability in different areas of the
state space, but we stick with Nosofsky's similarity function for the
time being, because it is a mathematically simple, yet established
notion in mathematical psychology.

@@

  \caption{Examples for Nosofsky-similarity (non-normalized) and
    derived probability of state confusion (normalized) for different
    values of $\alpha$ and $\card{\States} = 50$.}
  \label{fig:NosofskySim}
\end{figure}

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \alpha)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \alpha)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\alpha$ (see Figure~\ref{fig:NosofskySim}, bottom part). For $\alpha
= 0$ the confusion matrix has $C_{ii} = 1$; for $\alpha > 0$ $C$ is
positive, i.e., $C_{ij} >0$ for all $i$ and $j$; for $\alpha
\rightarrow \infty$ we find $C_{ij} = \nicefrac{1}{\card{\States}}$ for all
$i$ and $j$.

Difusion of behavior under confusion of states was defined in
Equation~(\ref{eq:confusion-function}), repeated here:
\begin{align*}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align*}
If states are confused on average with a probability proportional to
their similarity, the effect on noise-perturbed behavior is that
behavioral dispositions gradually diffuse along a gradient of
similarity of states as well. Consequently, iterated diffusion leads
to a smoothing out and an eventual equalization of behavioral
strategies, for both the sender and the
receiver. Figure~\ref{fig:confusion-SenRec} shows an arbitrary sender
and receiver strategy after one or several rounds of iterated
application of diffusion, for different values of perceptual
imprecision $\alpha$.

The example in Figure~\ref{fig:confusion-SenRec} shows that iterated
diffusion leads to sender and receiver behavior that is the same at
all choice points. After $n$ steps of diffusion an initial sender
strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial receiver
strategy will be $\Rstrat C^n$. If we assume that $\alpha > 0$ and so
$C$ is a positive row-stochastic matrix (each state could in principle
be confused as any other state, albeit with possibly a very low
probability), it follows from the Perron-Frobenius theorem that the
limit $C^\infty = \lim_{n \rightarrow \infty} C^n$ exists (it is the
Perron-projection) and all of its rows are identical. That is why, in
the limit of iterated diffusion, $C^\infty S$ has identical rows
(messages are sent with the same probability in each state), and $R
C^\infty$ likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

  \caption{Iterated diffusion for one arbitrary sender (top) and one
    receiver strategy (bottom), for a game with 50 states and 2
    messages. Each row corresponds to a different confusion level
    $\alpha$, each colum to a different number of applications of
    iterated diffusion.}
  \label{fig:confusion-SenRec}
\end{figure}

@@

@@

\subsubsection{Methodology}
\newcommand{\impairment}{\alpha}
In our simulations we focus on a concrete setup in line with \cite{Correia2013:The-Bivalent-Tr}.
Namely, we define a similarity-maximization game with $\States = \Acts$, where $\States$ is a discrete set of equidistant points within the unit interval $[0,1]$, including the boundaries (thus we always use $n \geq 2$).
As similarity metric $s:\States \times \States \rightarrow [0,1]$, we use a negative exponential of the of the square distance, \emph{i.e.}~for $\state,\state^\prime \in \States$:
\begin{align*}
  s(\state,\state^\prime) =
    \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \state = \state^\prime \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \state \neq \state^\prime \\
    e^{-\frac{\lVert \state - \state^\prime \rVert ^2}{\impairment^2}} & \textrm{otherwise} \\
    \end{cases}
\end{align*}
where $\impairment$ is an \emph{impairment} factor.
Here we deviate from \cite{Correia2013:The-Bivalent-Tr} by working with impairment $\impairment$ rather than acuity $\theta$, since we are interested in exploring the case of having $0$ impairment, and acuity does not have a natural maximum as defined.
Confusion is directly related with similarity, \emph{i.e.}~the more similar two states are, the more likely an agent is to perceive one as the other.
For simplicity, we equate one with the other by defining $C_{ij} = s(\state_i,\state_j)$.

>>>>

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \impairment) =
      \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^s}{ \impairment^2} \right ) \,,& \textrm{otherwise} \\
    \end{cases}
\end{align}
where $\impairment >= 0$ is an imprecision parameter. When $\impairment=0$ agents
perfectly discriminate between states; when $\impairment \rightarrow
\infty$ agents cannot discriminate states at
all. Figure~\ref{fig:NosofskySim} gives an impression of
Nosofsky-similarity for different parameter values. Other
formalizations of perceptual similarity are possible, including ones
that allow for different discriminability in different areas of the
state space, but we stick with Nosofsky's similarity function for the
time being, because it is a mathematically simple, yet established
notion in mathematical psychology.

@@

  \caption{Examples for Nosofsky-similarity (non-normalized) and
    derived probability of state confusion (normalized) for different
    values of $\impairment$ and $\card{\States} = 50$.}
  \label{fig:NosofskySim}
\end{figure}

We further assume that the probability of confusing any two states
$\mystate{i}$ and $\mystate{j}$ is proportional to their perceived
similarity and therefore obtained by normalization:
\begin{align*}
  C_{ij} = \frac{\similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} {\sum_j \similarity(\mystate{i}, \mystate{j} \myts ; \myts
  \impairment)} \,.
\end{align*}
Confusion of states is then also a function of imprecision parameter
$\impairment$ (see Figure~\ref{fig:NosofskySim}, bottom part). For $\impairment
= 0$ the confusion matrix has $C_{ii} = 1$; for $\impairment > 0$ $C$ is
positive, i.e., $C_{ij} >0$ for all $i$ and $j$; for $\impairment
\rightarrow \infty$ we find $C_{ij} = \nicefrac{1}{\card{\States}}$ for all
$i$ and $j$.

Difusion of behavior under confusion of states was defined in
Equation~(\ref{eq:confusion-function}), repeated here:
\begin{align*}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align*}
If states are confused on average with a probability proportional to
their similarity, the effect on noise-perturbed behavior is that
behavioral dispositions gradually diffuse along a gradient of
similarity of states as well. Consequently, iterated diffusion leads
to a smoothing out and an eventual equalization of behavioral
strategies, for both the sender and the
receiver. Figure~\ref{fig:confusion-SenRec} shows an arbitrary sender
and receiver strategy after one or several rounds of iterated
application of diffusion, for different values of perceptual
imprecision $\impairment$.

The example in Figure~\ref{fig:confusion-SenRec} shows that iterated
diffusion leads to sender and receiver behavior that is the same at
all choice points. After $n$ steps of diffusion an initial sender
strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial receiver
strategy will be $\Rstrat C^n$. If we assume that $\impairment > 0$ and so
$C$ is a positive row-stochastic matrix (each state could in principle
be confused as any other state, albeit with possibly a very low
probability), it follows from the Perron-Frobenius theorem that the
limit $C^\infty = \lim_{n \rightarrow \infty} C^n$ exists (it is the
Perron-projection) and all of its rows are identical. That is why, in
the limit of iterated diffusion, $C^\infty S$ has identical rows
(messages are sent with the same probability in each state), and $R
C^\infty$ likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

  \caption{Iterated diffusion for one arbitrary sender (top) and one
    receiver strategy (bottom), for a game with 50 states and 2
    messages. Each row corresponds to a different confusion level
    $\impairment$, each colum to a different number of applications of
    iterated diffusion.}
  \label{fig:confusion-SenRec}
\end{figure}

@@

\begin{itemize}
\item utility function: another Nosofsky-style function
\item get data for different set sizes, impairment and utility parameters
\item 
\end{itemize}

\newpage

@@

\subsubsection{Methodology}

In our simulations we focus on a concrete setup in line with \cite{Correia2013:The-Bivalent-Tr}.
Namely, we define a similarity-maximization game with $\States = \Acts$, where $\States$ is a discrete set of equidistant points within the unit interval $[0,1]$, including the boundaries (thus we always use $n \geq 2$).
As similarity metric $s:\States \times \States \rightarrow [0,1]$, we use a negative exponential of the of the square distance, \emph{i.e.}~for $\state,\state^\prime \in \States$:
\begin{align*}
  s(\state,\state^\prime) =
    \begin{cases}
    1 & \textrm{if } \impairment = 0 \textrm{ and } \state = \state^\prime \\
    0 & \textrm{if } \impairment = 0 \textrm{ and } \state \neq \state^\prime \\
    e^{-\frac{\lVert \state - \state^\prime \rVert ^2}{\impairment^2}} & \textrm{otherwise} \\
    \end{cases}
\end{align*}
where $\impairment$ is an \emph{impairment} factor.
Here we deviate from \cite{Correia2013:The-Bivalent-Tr} by working with impairment $\impairment$ rather than acuity $\theta$, since we are interested in exploring the case of having $0$ impairment, and acuity does not have a natural maximum as defined.
Confusion is directly related with similarity, \emph{i.e.}~the more similar two states are, the more likely an agent is to perceive one as the other.
For simplicity, we equate one with the other by defining $C_{ij} = s(\state_i,\state_j)$.
